---
title: "PromoterOntology vignette"
author: "Dunja"
date: 'Sys.Date()'
output: 
  rmarkdown::html_vignette:
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(225)
```

This Vignette will present functionality of `PromoterOntology` package. 

`PromoterOntology` package provides functionality to obtain promoters, annotate them with user specified features and after annotating a group of genes, do an over-enrichment analysis for promoter features. 

Here, we will go through each of the steps mentioned above. 

```{r load_promOnto, message=FALSE, warning=FALSE}
library(PromoterOntology)
library(dplyr)
```


## Obtain biomart promoter annotations and create promoter regions

To start the analysis, we will obtain gene annotations from `Biomart`. Gene annotations from `Biomart` are based on available RNAseq data that is not capturing promoters well. For more accurate promoter position, we will update `Biomart` promoter coordinates using CAGE data. This is an important step since all downstream analyses depend on accurate promoter coordinates.

In this Vignette we will use we will use hg19 gene annotations and CAGE data from FANTOM5 GM12878 cell line. 

```{r biomart promoters, message=FALSE, warning=FALSE}
library(biomaRt)
library(BSgenome.Hsapiens.UCSC.hg19) ## for promoter sequences

hg_mart <- useMart(host="grch37.ensembl.org",
                   biomart='ENSEMBL_MART_ENSEMBL',
                   dataset = "hsapiens_gene_ensembl")

biomart_promoters <- get_biomart_promoters(hg_mart)
### getting rid of scaffolds
biomart_promoters <- biomart_promoters[grep("chr", seqnames(biomart_promoters))]
```

CAGE data used in this Vignette is used in form of tagClusters, clustered CAGE start sites with information on dominant start position. TagCluster object is obtained from `CAGEr` package as a data frame. For the example in this vignette we included TagCluster object for GM12878 cell line.
TagClusters are overlapped with Biomart annotated promoters. In case there are multiple TagClusters overlapping a single promoter, we will keep only the most strongly expressed TagCluster. 
Created promoters are centered on dominant Transcription Start Site (dTSS) and extended by 500 nt from both sides. 

For the benefit of speed of Vignette rendering, we will subset promoters to 200 random promoters. 


```{r, warning=FALSE, message=FALSE}
# read in tag clusters from CAGE experiment
tc <- readRDS("~/Work/PromoterOnto/rds_files/tcGM12878.RDS")
tc <- makeGRangesFromDataFrame(tc, keep.extra.columns = TRUE)

cage_proms <- get_CAGE_promoters(tc_gr = tc, promoters = biomart_promoters)

cage_subset <- cage_proms[sample(1:length(cage_proms), 200)]
```


## Promoter annotation

After creating promoter regions centred on dominant TSS, we will obtain JASPAR PolII core promoter collection of PWMs *(data included in the package and can be found by calling "pwmList")* and perform sequence scan with these PWMs. 

In case of analysis in this vignette, we are working with human sample, so from JASPAR core promoter motives, we will omit **MTE** and **DPE** motives since they have been shown to be Drosophilla specific. 


```{r scan sequences with PolII collection, message=FALSE, warning=FALSE}
data("pwmList")
## pwmList contains 13 PWMs, since 2 of those are specific for Drosophilla, we will remove those
TF_names <- c("MTE", "INR", "GC-box", "CCAAT-box", "DPE", "BREu", "BREd", "DCE_S_I", "DCE_S_II",
              "DCE_S_III", "XCPE1", "TATA-box", "MED-1")

### remove Drosophila specific motives
pwmList <- pwmList[c(2:4, 6:13)]
TF_names <- TF_names[c(2:4, 6:13)]

prom_seq <- getSeq(Hsapiens, cage_subset)

pwm_prom <- lapply(pwmList, function(x) scan_promoters(seq = prom_seq, pwm = x, min_score = "80%" ))
names(pwm_prom) <- TF_names
```

`pwm_prom` is a list that contains results of sequence scan, where each sublist represents one PWM motif scan. Sublists are dataframes that contain information about hits position, its score and sequence motif itself. 

Sequence scans are known to produce many false positive results. To go around this problem we propose a method that extracts two coefficients from scan hits: **localisation precision** and **motive accuracy**. Hits that will have high values for both coefficients are considered to be true hits since they are correctly positioned and are very similar to PWM that was used to scan sequences. 

Localisation precision reports how likely it is that reported hit is found on a correct position. Based on distribution of all hits from all promoters, a probability of hit appearing at each position along promoter region was calculated.

Motive accuracy informs how similar identified motif is to the feature PWM.

For many core promoter motives a region in which they can be found has been identified. The package provides information on these PWM location constraints for PolII motives. These can be used to additionally filter out hits appearing outside the expected promoter region. 

For all remaining hits, we estimate an optimal threshold for the combined score of localisation precision and motive accuracy that can be further used to filter out false positives. 

Since in this example we have significantly reduced number of promoters we are working with, for some motives there will be too few hits to confidently calculate thresholds, so we will omit those. 

```{r evaluate hits, warning=FALSE}
library(inflection)
eval_threshold <- lapply(pwm_prom, find_threshold)
names(eval_threshold) <- TF_names
## filter all sequence scan hits that are below threshold

keep_me <- sapply(eval_threshold, function(x) !is.na(x))

pwm_prom <- mapply(function(x, y) filter_hits(all_hits = x, threshold = y), 
               x = pwm_prom[keep_me], y = eval_threshold[keep_me], SIMPLIFY = FALSE)

```


#### Create promoter matrix object

A promoter_matrix object is a dataset in which each row represents a promoter while columns are promoter features. 

```{r create_prom_matrix, message=FALSE, warning=FALSE}
prom_matrix <- matrix(0, nrow=length(cage_subset), ncol = length(pwm_prom))
colnames(prom_matrix) <- names(pwm_prom)
rownames(prom_matrix) <- 1:length(cage_subset)

for(i in 1:length(pwm_prom)){
  pwm_prom[[i]]$seqnames <- as.numeric(as.character(pwm_prom[[i]]$seqnames))
  prom_matrix[pwm_prom[[i]]$seqnames, i] <- 1
}

prom_matrix <- as.data.frame(prom_matrix)
prom_matrix$gene <- mcols(cage_subset)$external_gene_name
prom_matrix <- cbind(prom_matrix, mcols(cage_subset)[, c("ensembl_gene_id", "gene_biotype")])

```


#### Annotate with CpG and bidirectionality information 

Information about bidirectional genes and CpG islands for hg19 has been provided by the package. To annotate promoters with bidirectionality information we will overlap these two dayasets by their ensamble_gene_id. 

```{r annotate CpG and bidir, message=FALSE, warning=FALSE}
data(bidir)

prom_matrix$bidir <- 0
prom_matrix$bidir[prom_matrix$ensembl_gene_id %in% bidir$ENSEMBL] <- 1

### annotate CpG islands
data(cgi)

promsGM_cgi <- resize(cage_subset, 200, fix = "center")
promsGM_cgi <- resize(promsGM_cgi, 100, fix="start")
over <- findOverlaps(promsGM_cgi, cgi)

prom_matrix$cgi <- 0
prom_matrix$cgi[prom_matrix$promIndex %in% unique(queryHits(over))] <- 1
```


### Analysing overrepresentation

Having promoter matrix ready, we can start analysing our groups of genes for specific promoter features using promoter_onto function. 

To run an overrepresentation test, we need to have a subset of promoters for which we want to test if they have significantly enriched promoter features. 
In this example we will randomly subset 30 promoters
```{r run_prom_onto_and_plot,  fig.height=6, fig.width=6}

gene_sample <- prom_matrix$gene[sample(1:nrow(prom_matrix), 30)]

overrep <- promoter_onto(promoter_matrix = prom_matrix, 
                         sample = gene_sample, 
                         proms = cage_subset,
                         features = c(1:6, 10:11),
                         label = "Vignette example", 
                         all = TRUE, save_plot = FALSE)

overrep
```


### Comparing multiple categories 

```{r plot_multiple_POs,  fig.height=5, fig.width=5}

## create 4 random groups of genes
gene_sample_list <- list()
for(i in 1:4){
  gene_sample_list[[i]] <- prom_matrix$gene[sample(1:nrow(prom_matrix), 30)]
}

overrep_list <- lapply(gene_sample_list, function(x) promoter_onto(promoter_matrix = prom_matrix, 
                         sample = x, 
                         proms = cage_subset,
                         features = c(1:6, 10:11), 
                         all = TRUE, save_plot = TRUE, 
                         label="multiple groups"))
names(overrep_list) <- paste0("group_", 1:4)

plot_multiple_POs(overrep_list) + ggplot2::ggtitle("Overenrichment of promoter features for \n 4 randomly created gene groups")
```
